Effect of Black Hole Spins on Christodoulou Memory in Binary Mergers and its implication on Observing Gravitational Wave Bursts With Memory in Pulsar Timing Array

### 2019-10-09 notes/Paper.tex
updated paper

\par The corrections to the leading-oder formula for the nonlinear memory's contribution to 3PN order for non-spinning binaries has been computed by by Favata \cite{Favata2009}. However the effect of binary spins have not been explored in these results. Astrophysical black holes are formed by gravitation collapse of massive star when when they run out of their nuclear fuel in the core. These massive stars are likely to posses spin angular momentum, as a result the black hole formed would also be spinning. The black holes can also accrete matter from surrounding and as a result spin up during the process. It is highly likely that there is a large population of spinning binaries in the universe.      
\par In this paper we calculate the nonlinear memory contribution to the + waveform polarization and look at the effect of spin contribution. For PN calculations, this paper is just an extension of Favata's\cite{Favata2009} paper, where we include the spin contribution up to leading (1.5 PN) order for nonprecessing binaries. The nonlinear memory contribution near merger is computed using NR data from publicly available SXS catalog, where we process the data to remove the low frequency numerical noise. In Sec. II, we describe the calculation required to compute memory contribution to the post-Newtonian waveform of quasi-circular, inspiralling binaries. In Sec. III, we describe the method used to extract memory contribution from NR simulations. Sec. IV we give the results which looks at the effect of spin on total nonlinear memory. We use PN expressions for memory for early time and for late time we use memory extracted from NR simulations.

@@

\begin{figure}
	\includegraphics[width=3.5in]{../plots/PlotfromMathematicaData/MemoryResSpinvsMass14Days.pdf}
	\includegraphics[width=3.5in]{../plots/PlotfromMathematicaData/MemoryResSpinvsMass1825Days.pdf}
	\caption{The top figure shows the root mean square of post-fit timing residuals growth as a function of Mass and Spin magnitude $\chi$. The left plot shows this post-fit residual fit during two weeks interval of time and right plot shows this for 5 years interval.}
	\label{fig:MemoryMassVsSpin}
\end{figure}
\end{widetext}
\section{CONCLUSIONS}
Observation of Christodoulou memory would be an important test of general relativity in strong field regime. Effort to detect this memory effect in low frequency band from super massive black holes have been undertaken by pulsar timing array experiments \cite{Arzoumanian2015}. The author in their analysis have looked for memory in a observation data for 5 years. In their analysis they have assumed the memory to a step function for a given amplitude. This amplitude is of the order $10^{-15}$, which is memory jump for a $10^{8}M_{\odot}$ binary. In these analysis the spin effect have been neglected for modeling the signal. It is highly likely that super massive black holes have significant spin due to accretion of matter from surrounding. We have seen that the spin can significantly change memory amplitude as a result it can change the timing residuals. In our analysis we observe that:(1) Growth of memory when looks over a shorter observing time scale does not look like a step function, instead it's shape depends upon the spin and mass of binary for a give interval of time.(2) The post-fit residual for a observation period of two weeks  .Hence it is important to consider the spin effect while modeling the signals. The important thing for pulsar timing measurements would be the amount of memory changes over a single two weeks period of time which is shown the figure \ref{fig:MemoryMassVsSpin} as function of total mass of the binary and the spin magnitude. For a given mass and over the allowed spin range the net change in the memory amplitude can be as big as a factor of 5 which would change the SNR, or the expected event rate within a given volume, by a factor of $5^{3}=125$

>>>>

\par The corrections to the leading-oder formula for the nonlinear memory's contribution to 3PN order for non-spinning binaries has been computed by by Favata \cite{Favata2009}. However the effect of binary spins have not been explored in these results. Astrophysical black holes are formed by gravitation collapse of massive star when when they run out of their nuclear fuel in the core. These massive stars are likely to posses spin angular momentum, as a result the black hole formed would also be spinning. The black holes can also accrete matter from surrounding and as a result spin up during the process. It is highly likely that there is a large population of spinning binaries in the universe.      
\par In this paper we calculate the nonlinear memory contribution to the + waveform polarization and look at the effect of spin contribution. For PN calculations, this paper is just an extension of Favata's\cite{Favata2009} paper, where we include the spin contribution up to leading (1.5 PN) order for nonprecessing binaries. The nonlinear memory contribution near merger is computed using NR data from publicly available SXS catalog, where we process the data to remove the low frequency numerical noise. In Sec. II, we describe the calculation required to compute memory contribution to the post-Newtonian waveform of quasi-circular, inspiralling binaries. In Sec. III, we describe the method used to extract memory contribution from NR simulations. Sec. IV we give the results which looks at the effect of spin on total nonlinear memory. We use PN expressions for memory for early time and for late time we use memory extracted from NR simulations. In Sec. V we look at how these results can effect expected results in effort to detect the BWM from super massive binaries in pulsar timing array. We look a the timing residuals growth over a period of two weeks interval of time.

@@

\begin{figure}
	\includegraphics[width=3.5in]{../plots/PlotfromMathematicaData/MemoryResSpinvsMass14Days.pdf}
	\includegraphics[width=3.5in]{../plots/PlotfromMathematicaData/MemoryResSpinvsMass1825Days.pdf}
	\caption{The top figure shows the root mean square of post-fit timing residuals growth as a function of Mass and Spin magnitude $\chi$. The left plot shows this post-fit residual fit during two weeks interval of time and right plot shows this for 5 years interval.}
	\label{fig:MemoryMassVsSpin}
\end{figure}
\end{widetext}
\section{CONCLUSIONS}
Observation of Christodoulou memory would be an important test of general relativity in strong field regime. Effort to detect this memory effect in low frequency band from super massive black holes have been undertaken by pulsar timing array experiments \cite{Arzoumanian2015}. The author in their analysis have looked for memory in a observation data for 5 years. In their analysis they have assumed the memory to a step function for a given amplitude. This amplitude is of the order $10^{-15}$, which is memory jump for a $10^{8}M_{\odot}$ binary. In these analysis the spin effect have been neglected for modeling the signal. It is highly likely that super massive black holes have significant spin due to accretion of matter from surrounding. We have seen that the spin can significantly change memory amplitude as a result it can change the timing residuals. In our analysis we observe that:(1) Growth of memory when looked over a shorter observing time scale does not look like a step function, instead it's shape depends upon the spin and mass of binary for a give interval of time.(2) The post-fit rms residual for a observation period of two weeks reach a maximum for $10^{9}M_{\odot}$ and tend to decrease higher mass case. Hence it is important to consider the spin effect while modeling the BWM signals and consider a range of mass values. Another important thing for pulsar timing measurements would be the amount of memory changes over a single two weeks period of time which is shown the figure \ref{fig:MemoryMassVsSpin} as function of total mass of the binary and the spin magnitude. For a given mass and over the allowed spin range the net change in the memory amplitude can be as big as a factor of 5 which would change the SNR, or the expected event rate within a given volume, by a factor of $5^{3}=125$
